---
title: "Basic usage of tidycensus"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 8)
```

To get started working with __tidycensus__, users should load the package along with the __tidyverse__ package, and set their Census API key.  A key can be obtained from <http://api.census.gov/data/key_signup.html>.  

```{r, echo = FALSE}
library(tidycensus)
library(tidyverse)

census_api_key("5ed58a5745802102fb83d4eec5d1f7326f65ffab")
```

```{r, eval = FALSE}
library(tidycensus)
library(tidyverse)

census_api_key("YOUR API KEY GOES HERE")
```

There are two major functions implemented in __tidycensus__: `get_decennial`, which grants access to the 1990, 2000, and 2010 decennial US Census APIs, and `get_acs`, which grants access to the 5-year American Community Survey APIs.  In this basic example, let's look at median gross rent by state in 1990: 

```{r}
m90 <- get_decennial(geography = "state", variables = "H043A001", year = 1990)

head(m90)
```

The function returns a tibble with four columns by default: `GEOID`, which is an identifier for the geographical unit associated with the row; `NAME`, which is a descriptive name of the geographical unit; `variable`, which is the Census variable represented in the row; and `value`, which is the value of the variable for that unit.  By default, __tidycensus__ functions return tidy data frames in which rows represent unit-variable combinations; for a wide data frame with Census variable names in the columns, set `output = "wide"` in the function call.  

As the function has returned a tidy object, we can visualize it quickly with __ggplot2__: 

```{r, fig.height = 8}
m90 %>%
  ggplot(aes(x = value, y = reorder(NAME, value))) + 
  geom_point()
```

## Searching for variables

Getting variables from the Census or ACS requires knowing the variable ID - and there are thousands of these IDs across the different Census files.  To rapidly search for variables, use the `load_variables` function.  The function takes two required arguments: the year of the Census or endyear of the ACS sample, and the dataset - one of `"sf1"`, `"sf3"`, or `"acs5"`. For ideal functionality, I recommend assigning the result of this function to a variable, setting `cache = TRUE` to store the result on your computer for future access, and using the `View` function in RStudio to interactively browse for variables.  

```{r, eval = FALSE}
v15 <- load_variables(2016, "acs5", cache = TRUE)

View(v15)
```

<img src=img/view.png style="width: 100%">

By filtering for "median age" I can quickly view the variable IDs that correspond to my query.  

## Working with ACS data

American Community Survey data differ from decennial Census data in that ACS data are based on an annual sample of approximately 3 million households, rather than a more complete enumeration of the US population.  In turn, ACS data points are __estimates__ characterized by a __margin of error__.  __tidycensus__ will always return the estimate and margin of error together for any requested variables.  In turn, when requesting ACS data with __tidycensus__, it is not necessary to specify the `"E"` or `"M"` suffix for a variable name.  Let's fetch median household income data from the 2011-2015 ACS for counties in Vermont; the endyear is not necessary here as the function defaults to 2015. 

```{r}
vt <- get_acs(geography = "county", 
              variables = c(medincome = "B19013_001"), 
              state = "VT")

vt
```

The output is similar to a call to `get_decennial`, but instead of a `value` column, `get_acs` returns `estimate` and `moe` columns for the ACS estimate and margin of error, respectively.  `moe` represents the default 90 percent confidence level around the estimate; this can be changed to 95 or 99 percent with the `moe_level` parameter in `get_acs` if desired. 

As we have the margin of error, we can visualize the uncertainty around the estimate: 

```{r}
vt %>%
  mutate(NAME = gsub(" County, Vermont", "", NAME)) %>%
  ggplot(aes(x = estimate, y = reorder(NAME, estimate))) +
  geom_errorbarh(aes(xmin = estimate - moe, xmax = estimate + moe)) +
  geom_point(color = "red", size = 3) +
  labs(title = "Household income by county in Vermont",
       subtitle = "2012-2016 American Community Survey",
       y = "",
       x = "ACS estimate (bars represent margin of error)")
```
