---
title: "Basic usage of tidycensus"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 8)
```

To get started working with __tidycensus__, users should load the package along with the __tidyverse__ package, and set their Census API key.  A key can be obtained from <http://api.census.gov/data/key_signup.html>.  

```{r, echo = FALSE}
library(tidycensus)
library(tidyverse)
```

```{r, eval = FALSE}
library(tidycensus)
library(tidyverse)

census_api_key("YOUR API KEY GOES HERE")
```

There are two major functions implemented in __tidycensus__: `get_decennial()`, which grants access to the 2000 and 2010 decennial US Census APIs, and `get_acs()`, which grants access to the 1-year and 5-year American Community Survey APIs.  _Note: at the moment, access to the 1990 (SF1 and SF3) and 2000 (SF3) APIs has been suspended._ 

In this basic example, let's look at median age by state in 2010: 

```{r}
age10 <- get_decennial(geography = "state", 
                       variables = "P013001", 
                       year = 2010)

head(age10)
```

The function returns a tibble with four columns by default: `GEOID`, which is an identifier for the geographical unit associated with the row; `NAME`, which is a descriptive name of the geographical unit; `variable`, which is the Census variable represented in the row; and `value`, which is the value of the variable for that unit.  By default, __tidycensus__ functions return tidy data frames in which rows represent unit-variable combinations; for a wide data frame with Census variable names in the columns, set `output = "wide"` in the function call.  

As the function has returned a tidy object, we can visualize it quickly with __ggplot2__: 

```{r, fig.height = 8}
age10 %>%
  ggplot(aes(x = value, y = reorder(NAME, value))) + 
  geom_point()
```

## Geography in tidycensus

To get decennial Census data or American Community Survey data, tidycensus users supply an argument to the required `geography` parameter.  Arguments are formatted as consumed by the Census API, and specified in the table below.  Not all geographies are available for all surveys, all years, and all variables.  Most Census geographies are supported in tidycensus at the moment; if you require a geography that is missing from the table below, please file an issue at https://github.com/walkerke/tidycensus/issues.  

If __state__ or __county__ is in bold face in "Available by", you are required to supply a state and/or county for the given geography.  

|Geography|Definition|Available by|Available in|
|---------|----------|------------|------------|
|`"us"`|United States||`get_acs()`, `get_decennial()`| 
|`"region"`|Census region||`get_acs()`, `get_decennial()`| 
|`"division"`|Census division||`get_acs()`, `get_decennial()`| 
|`"state"`|State or equivalent|state|`get_acs()`, `get_decennial()`| 
|`"county"`|County or equivalent|state, county|`get_acs()`, `get_decennial()`| 
|`"county subdivision"`|County subdivision|__state__, county|`get_acs()`, `get_decennial()`| 
|`"tract"`|Census tract|__state__, county|`get_acs()`, `get_decennial()`|  
|`"block group"`|Census block group|__state__, county|`get_acs()`, `get_decennial()`|  
|`"block"`|Census block|__state__, __county__|`get_decennial()`| 
|`"place"`|Census-designated place|state|`get_acs()`, `get_decennial()`|  
|`"alaska native regional corporation"`|Alaska native regional corporation|state|`get_acs()`, `get_decennial()`|  
|`"american indian area/alaska native area/hawaiian home land"`|Federal and state-recognized American Indian reservations and Hawaiian home lands|state|`get_acs()`, `get_decennial()`|  
|`"american indian area/alaska native area (reservation or statistical entity only)"`|Only reservations and statistical entities|state|`get_acs()`, `get_decennial()`| 
|`"american indian area (off-reservation trust land only)/hawaiian home land"`|Only off-reservation trust lands and Hawaiian home lands|state|`get_acs()`| 
|`"metropolitan statistical area/micropolitan statistical area"`|Core-based statistical area|state|`get_acs()`, `get_decennial()`|
|`"combined statistical area"`|Combined statistical area|state|`get_acs()`, `get_decennial()`| 
|`"new england city and town area"`|New England city/town area|state|`get_acs()`, `get_decennial()`| 
|`"combined new england city and town area"`|Combined New England area|state|`get_acs()`, `get_decennial()`| 
|`"urban area"`|Census-defined urbanized areas||`get_acs()`, `get_decennial()`| 
|`"congressional district"`|Congressional district for the year-appropriate Congress|state|`get_acs()`, `get_decennial()`|  
|`"school district (elementary)"`|Elementary school district|__state__|`get_acs()`, `get_decennial()`| 
|`"school district (secondary)"`|Secondary school district|__state__|`get_acs()`, `get_decennial()`| 
|`"school district (unified)"`|Unified school district|__state__|`get_acs()`, `get_decennial()`| 
|`"public use microdata area"`|PUMA (geography associated with Census microdata samples)|state|`get_acs()`| 
|`"zip code tabulation area"` OR `"zcta"`|Zip code tabulation area||`get_acs()`, `get_decennial()`|  
|`"state legislative district (upper chamber)"`|State senate districts|__state__|`get_acs()`, `get_decennial()`|  
|`"state legislative district (lower chamber)"`|State house districts|__state__|`get_acs()`, `get_decennial()`|  


## Searching for variables

Getting variables from the Census or ACS requires knowing the variable ID - and there are thousands of these IDs across the different Census files.  To rapidly search for variables, use the `load_variables()` function.  The function takes two required arguments: the year of the Census or endyear of the ACS sample, and the dataset - one of `"sf1"`, `"sf3"`, or `"acs5"`. For ideal functionality, I recommend assigning the result of this function to a variable, setting `cache = TRUE` to store the result on your computer for future access, and using the `View` function in RStudio to interactively browse for variables.  

```{r, eval = FALSE}
v17 <- load_variables(2017, "acs5", cache = TRUE)

View(v17)
```

<img src=img/view.png style="width: 100%">

By filtering for "median age" I can quickly view the variable IDs that correspond to my query.  

## Working with ACS data

American Community Survey data differ from decennial Census data in that ACS data are based on an annual sample of approximately 3 million households, rather than a more complete enumeration of the US population.  In turn, ACS data points are __estimates__ characterized by a __margin of error__.  __tidycensus__ will always return the estimate and margin of error together for any requested variables when using `get_acs()`.  In turn, when requesting ACS data with __tidycensus__, it is not necessary to specify the `"E"` or `"M"` suffix for a variable name.  Let's fetch median household income data from the 2014-2018 ACS for counties in Vermont. 

```{r}
vt <- get_acs(geography = "county", 
              variables = c(medincome = "B19013_001"), 
              state = "VT", 
              year = 2018)

vt
```

The output is similar to a call to `get_decennial()`, but instead of a `value` column, `get_acs` returns `estimate` and `moe` columns for the ACS estimate and margin of error, respectively.  `moe` represents the default 90 percent confidence level around the estimate; this can be changed to 95 or 99 percent with the `moe_level` parameter in `get_acs` if desired. 

As we have the margin of error, we can visualize the uncertainty around the estimate: 

```{r}
vt %>%
  mutate(NAME = gsub(" County, Vermont", "", NAME)) %>%
  ggplot(aes(x = estimate, y = reorder(NAME, estimate))) +
  geom_errorbarh(aes(xmin = estimate - moe, xmax = estimate + moe)) +
  geom_point(color = "red", size = 3) +
  labs(title = "Household income by county in Vermont",
       subtitle = "2014-2018 American Community Survey",
       y = "",
       x = "ACS estimate (bars represent margin of error)")
```
