---
title: "Margins of error in the ACS"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidycensus)
library(tidyverse)
options(tibble.print.min = 6)
```

As opposed to the decennial Census datasets available with __tidycensus__, datasets from the five-year American Community Survey include _estimates_ with an associated _margin of error_, as the ACS is based on an annual sample of around 3 million households.  As explained by the Census Bureau: 

> While the main function of the decennial census is to
provide _counts_ of people for the purpose of congressional
apportionment and legislative redistricting, the
primary purpose of the ACS is to measure the changing
social and economic _characteristics_ of the U.S. population.
As a result, the ACS does not provide official
counts of the population in between censuses. [^1]

In turn, ACS data __do not__ represent precise counts of population subgroups, but are rather designed to give a general sense of how socioeconomic indicators vary across the country.  In many cases, ACS margins of error can be quite large - at times exceeding the estimate.  Let's say we want to study aging populations in Ramsey County, Minnesota from the 2012-2016 ACS by Census tracts.  We can pull up a dataset with `get_acs`: 

```{r}
library(tidycensus)
library(tidyverse)

vars <- paste0("B01001_0", c(20:25, 44:49))

ramsey <- get_acs(geography = "tract", 
                  variables = vars, 
                  state = "MN", 
                  county = "Ramsey", 
                  year = 2016)

head(ramsey %>% select(-NAME))
```

We can see that in two instances the margin of error exceeds the estimate.  One way to address this is through data aggregation.  While the specific group estimates in this Census tract may be unreliable, the estimate for the total population over age 65 in the tract is likely better, but is not available directly from the ACS detailed tables.  In turn, we can summarize our data.  

The US Census Bureau publishes guidelines on how to calculate margins of error for derived estimates (see the footnotes). These formulas are implemented in __tidycensus__ in the `moe_sum`, `moe_prop`, `moe_ratio`, and `moe_product` functions. The example below illustrates the use of `moe_sum` to calculate the margin of error around a derived estimate for Census tract population over age 65. 

```{r}
ramsey65 <- ramsey %>%
  group_by(GEOID) %>%
  summarize(sumest = sum(estimate), 
            summoe = moe_sum(moe, estimate))

head(ramsey65)
```

The margins of error for this aggregate population are more reasonable.  However, the Census Bureau does issue this warning: 

> These methods do not consider the correlation or covariance between the basic
estimates. They may be overestimates or underestimates of the derived estimate’s standard error,
depending on whether the two basic estimates are highly correlated in either the positive or
negative direction. As a result, the approximated standard error may not match direct
calculations of standard errors or calculations obtained through other methods. [^2]

Dealing with margins of error in the ACS is a complex, yet important topic.  This is especially true when dealing with spatial data, and considering ways to aggregate areal units to improve the reliability of estimates.  I recommend the following papers for further reading: 

Spielman, S.E., and Folch, D.C. (2015).  Reducing Uncertainty in the American Community Survey through Data-Driven Regionalization. _PLOS ONE_. http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0115626; Python implementation at https://github.com/geoss/ACS_Regionalization

Spielman, S.E., and Singleton, A. (2015) Studying Neighborhoods Using Uncertain Data from the American Community Survey: A Contextual Approach. _Annals of the Association of American Geographers_. http://www.tandfonline.com/doi/full/10.1080/00045608.2015.1052335; R implementation at https://github.com/geoss/acs_demographic_clusters.  

Wong, D.W., and Sun, M. (2013). Handling Data Quality Information of Survey Data in GIS: A Case of Using the American Community Survey Data. _Spatial Demography_. https://link.springer.com/article/10.1007/BF03354884. 

[^1]: United States Census Bureau (2008). A Compass for Understanding and Using American Community Survey Data.  https://www.census.gov/content/dam/Census/library/publications/2008/acs/ACSGeneralHandbook.pdf.  
[^2]: United States Census Bureau (2016). Instructions for Applying Statistical Testing to the 2011-2015 ACS 5-Year Data. https://www2.census.gov/programs-surveys/acs/tech_docs/statistical_testing/2015StatisticalTesting5year.pdf.  
