---
title: "Spatial data in tidycensus"
output: html_document
---


```{r setup, include=FALSE}
library(tidycensus)
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 8)
census_api_key("5ed58a5745802102fb83d4eec5d1f7326f65ffab")
```

If requested, __tidycensus__ can return simple feature geometry for geographic units along with variables from the decennial US Census or American Community survey.  By setting `geometry = TRUE` in a __tidycensus__ function call, __tidycensus__ will use the __tigris__ package to retrieve the corresponding geographic dataset from the US Census Bureau and pre-merge it with the tabular data obtained from the Census API.  The following example shows median household income from the 2011-2015 ACS for Census tracts in Orange County, California: 

```{r}
library(tidycensus)
library(tidyverse)
options(tigris_use_cache = TRUE)

orange <- get_acs(state = "CA", county = "Orange", geography = "tract", 
                  variables = "B19013_001", geometry = TRUE)

head(orange)
```

Our object `orange` looks much like the basic __tidycensus__ output, but with a `geometry` list-column describing the geometry of each feature, using the geographic coordinate system NAD 1983 (EPSG: 4269) which is the default for Census shapefiles.  __tidycensus__ uses the Census [cartographic boundary shapefiles](https://www.census.gov/geo/maps-data/data/tiger-cart-boundary.html) for faster processing; if you prefer the TIGER/Line shapefiles, set `cb = FALSE` in the function call. 

As the dataset is in a tidy format, it can be quickly visualized with the `geom_sf` functionality currently in the development version of __ggplot2__: 

```{r}
library(viridis)

orange %>%
  ggplot(aes(fill = estimate, color = estimate)) + 
  geom_sf() + 
  coord_sf(crs = 26911) + 
  scale_fill_viridis(option = "magma") + 
  scale_color_viridis(option = "magma")
```

## Faceted mapping

One of the most powerful features of __ggplot2__ is its support for small multiples, which works very well with the tidy data format returned by __tidycensus__.  Many Census and ACS variables return _counts_, however, which are generally inappropriate for choropleth mapping.  In turn, `get_decennial` and `get_acs` have an optional argument, `summary_var`, that can work as a multi-group denominator when appropriate.  Let's use the following example of the racial geography of Harris County, Texas.  First, we'll request data for non-Hispanic whites, non-Hispanic blacks, non-Hispanic Asians, and Hispanics by Census tract for the 2010 Census, and specify total population as the summary variable.  `year` is not necessary here as the default is 2010.   

```{r}
racevars <- c(White = "P0050003", 
              Black = "P0050004", 
              Asian = "P0050006", 
              Hispanic = "P0040003")

harris <- get_decennial(geography = "tract", variables = racevars, 
                  state = "TX", county = "Harris County", geometry = TRUE,
                  summary_var = "P0010001") 

head(harris)
```

We notice that there are four entries for each Census tract, with each entry representing one of our requested variables. The `summary_value` column represents the value of the summary variable, which is total population in this instance.  When a summary variable is specified in `get_acs`, both `summary_est` and `summary_moe` columns will be returned.  

With this information, we can set up an analysis pipeline in which we calculate a new percent-of-total column; recode the Census variable names into more intuitive labels; and visualize the result for each group in a faceted plot.  

```{r}
harris %>%
  mutate(pct = 100 * (value / summary_value)) %>%
  ggplot(aes(fill = pct, color = pct)) +
  facet_wrap(~variable) +
  geom_sf() +
  coord_sf(crs = 26915) + 
  scale_fill_viridis() +
  scale_color_viridis()
```

Beyond this, you might be interested in writing your dataset to a shapefile or GeoJSON for use in external GIS or visualization applications.  You can accomplish this with the `st_write` function in the __sf__ package: 

```{r, eval = FALSE}
library(sf)
st_write(orange, "orange.shp")
```

Your __tidycensus__-obtained dataset can now be used in ArcGIS, QGIS, Tableau, or any other application that reads shapefiles.  

There is a lot more you can do with the spatial functionality in __tidycensus__, including more sophisticated visualization and spatial analysis; look for updates [on my blog](https://walkerke.github.io) and in this space.             
  
  