---
title: "Spatial data in tidycensus"
output: html_document
---


```{r setup, include=FALSE}
library(tidycensus)
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 8)
```

If requested, __tidycensus__ can return simple feature geometry for geographic units along with variables from the decennial US Census or American Community survey.  By setting `geometry = TRUE` in a __tidycensus__ function call, __tidycensus__ will use the __tigris__ package to retrieve the corresponding geographic dataset from the US Census Bureau and pre-merge it with the tabular data obtained from the Census API.  As of tidycensus version 0.9.9.2, `geometry = TRUE` is supported for all geographies currently available in the package.  

The following example shows median household income from the 2015-2019 ACS for Census tracts in Orange County, California: 

```{r}
library(tidycensus)
library(tidyverse)
options(tigris_use_cache = TRUE)

orange <- get_acs(state = "CA", county = "Orange", geography = "tract", 
                  variables = "B19013_001", geometry = TRUE)

head(orange)
```

Our object `orange` looks much like the basic __tidycensus__ output, but with a `geometry` list-column describing the geometry of each feature, using the geographic coordinate system NAD 1983 (EPSG: 4269) which is the default for Census shapefiles.  __tidycensus__ uses the Census [cartographic boundary shapefiles](https://www.census.gov/geographies/mapping-files/time-series/geo/carto-boundary-file.html) for faster processing; if you prefer the TIGER/Line shapefiles, set `cb = FALSE` in the function call. 

As the dataset is in a tidy format, it can be quickly visualized with the `geom_sf` functionality currently in the development version of __ggplot2__: 

```{r}
orange %>%
  ggplot(aes(fill = estimate)) + 
  geom_sf(color = NA) + 
  coord_sf(crs = 26911) + 
  scale_fill_viridis_c(option = "magma") 
```

Please note that the UTM Zone 11N coordinate system (`26911`) is appropriate for Southern California but may not be for your area of interest.  For help identifying an appropriate projected coordinate system for your data, take a look at [the {crsuggest} R package](https://github.com/walkerke/crsuggest).

## Faceted mapping

One of the most powerful features of __ggplot2__ is its support for small multiples, which works very well with the tidy data format returned by __tidycensus__.  Many Census and ACS variables return _counts_, however, which are generally inappropriate for choropleth mapping.  In turn, `get_decennial` and `get_acs` have an optional argument, `summary_var`, that can work as a multi-group denominator when appropriate.  Let's use the following example of the racial geography of Harris County, Texas.  First, we'll request data for non-Hispanic whites, non-Hispanic blacks, non-Hispanic Asians, and Hispanics by Census tract for the 2010 Census, and specify total population as the summary variable.  `year` is not necessary here as the default is 2010.   

```{r}
racevars <- c(White = "P005003", 
              Black = "P005004", 
              Asian = "P005006", 
              Hispanic = "P004003")

harris <- get_decennial(geography = "tract", variables = racevars, 
                  state = "TX", county = "Harris County", geometry = TRUE,
                  summary_var = "P001001") 

head(harris)
```

We notice that there are four entries for each Census tract, with each entry representing one of our requested variables. The `summary_value` column represents the value of the summary variable, which is total population in this instance.  When a summary variable is specified in `get_acs`, both `summary_est` and `summary_moe` columns will be returned.  

With this information, we can set up an analysis pipeline in which we calculate a new percent-of-total column; recode the Census variable names into more intuitive labels; and visualize the result for each group in a faceted plot.  

```{r}
harris %>%
  mutate(pct = 100 * (value / summary_value)) %>%
  ggplot(aes(fill = pct)) +
  facet_wrap(~variable) +
  geom_sf(color = NA) +
  coord_sf(crs = 26915) + # Appropriate CRS for Houston, but may not be for your data
  scale_fill_viridis_c()
```

## Detailed shoreline mapping with tidycensus

Geometries in tidycensus default to the Census Bureau's [cartographic boundary shapefiles](https://www.census.gov/geo/maps-data/data/tiger-cart-boundary.html).  Cartographic boundary shapefiles are preferred to the core [TIGER/Line shapefiles](https://www.census.gov/geo/maps-data/data/tiger-line.html) in tidycensus as their smaller size speeds up processing and because they are pre-clipped to the US coastline.

However, there may be circumstances in which your mapping requires more detail.  A good example of this would be maps of New York City, in which even the cartographic boundary shapefiles include water area.  For example, take this example of median household income by Census tract in Manhattan (New York County), NY: 

```{r, message = FALSE, warning = FALSE}
library(tidycensus)
library(mapview)
options(tigris_use_cache = TRUE)

ny <- get_acs(geography = "tract", 
              variables = "B19013_001", 
              state = "NY", 
              county = "New York", 
              geometry = TRUE)

mapview(ny, zcol = "estimate", legend = TRUE)

```

As illustrated in the graphic, the boundaries of Manhattan include water boundaries - stretching into the Hudson and East Rivers.  In turn, a more accurate representation of Manhattan's land area might be desired.  To accomplish this, a tidycensus user can use the core TIGER/Line shapefiles instead, then erase water area from Manhattan's geometry.  

tidycensus allows users to get TIGER/Line instead of cartographic boundary shapefiles with the keyword argument `cb = FALSE`.  This argument will be familiar to users of the tigris package, as it is used by tigris to distinguish between cartographic boundary and TIGER/Line shapefiles in the package.  

```{r, message = FALSE, warning = FALSE}
library(sf)

ny2 <- get_acs(geography = "tract", 
              variables = "B19013_001", 
              state = "NY", 
              county = "New York", 
              geometry = TRUE, 
              cb = FALSE) 
```

Next, tools in the tigris and [sf](https://github.com/r-spatial/sf) package can be used to remove the water area from Manhattan's Census tracts.  sf allows users to "erase" one geometry from another, [akin to tools available in desktop GIS software](http://pro.arcgis.com/en/pro-app/tool-reference/analysis/erase.htm).  The `st_erase()` function defined below is not exported by the package, but is defined in the documentation for `st_difference()`.  

The geometry used to "erase" water area from the tract polygons is obtained by the `area_water()` function in tigris, making sure to choose the option `class = "sf"`.  

```{r, message = FALSE, warning = FALSE}
library(tigris)

st_erase <- function(x, y) {
  st_difference(x, st_union(y))
}

ny_water <- area_water("NY", "New York", class = "sf") 

ny_erase <- st_erase(ny2, ny_water)

```

After performing this operation, we can visualize the result: 

```{r, message = FALSE, warning = FALSE}
mapview(ny_erase, zcol = "estimate", legend = TRUE)
```

The map appears as before, but instead the polygons now hug the shoreline of Manhattan.  


## Writing to shapefiles

Beyond this, you might be interested in writing your dataset to a shapefile or GeoJSON for use in external GIS or visualization applications.  You can accomplish this with the `st_write` function in the __sf__ package: 

```{r, eval = FALSE}
library(sf)
st_write(orange, "orange.shp")
```

Your __tidycensus__-obtained dataset can now be used in ArcGIS, QGIS, Tableau, or any other application that reads shapefiles.  

There is a lot more you can do with the spatial functionality in __tidycensus__, including more sophisticated visualization and spatial analysis; look for updates [on my blog](https://walkerke.github.io) and in this space.             
  
  
